Quantifying Stability and Change in Personal Culture Using Panel Data

### 2023-09-21 manuscript.Rmd

---
title: "Quantifying Stability and Change in Personal Culture Using Panel Data"
author:
- 
output:
  pdf_document:
    keep_tex: true
bibliography: kileybib.bib
csl: asa.csl
abstract: "Recent work has produced conflicting interpretations on the question of whether adults undergo durable changes in their personal culture over time. This paper asserts that this divergence is due more to the approach adopted by researchers –-a tournament of models-– than their findings, which are highly consonant. We reanalyze the data used previously in this debate with an approach that quantifies the processes of cultural difference, rather than one that detects the presence or absence of change. In doing so, we compare how much cultural difference is explained by intrapersonal change and interpersonal differences. Our results harmonize recent findings by showing that, while most measures of personal culture show change over time, the variance explained by intrapersonal change is typically small compared to interpersonal differences at baseline. We then exemplify how quantifying these processes offer new ways to address theoretical questions of when and how the relative importance of intrapersonal change and interpersonal differences shifts."
fontsize: 11pt
---

@@

A recurring focus in the sociology of culture is the question of whether people, as they move through their lives and transition between social roles, change their personal culture –- the declarative and non-declarative attitudes, beliefs, values, and dispositions that manifest at the individual level [@lizardo2017; @kiley2020; @lersch2023].

This seemingly simple question underlies a number of central debates. For example, pragmatist theories of action claim that changes in social environments cause people to adapt their views and make new meanings [@swidler2001; @gross2009], while Bourdiesian practice theories argue that the “past conditions of production” leave a lasting mark on peoples’ personal culture throughout their lives [@bourdieu1990]. Similarly, models of social influence require people to adapt their culture in the face of new information [@vaisey2010a], while models of aggregate social and cultural change that place emphasis on cohort effects require them to be open to change while young but fairly resistant to change as they age [@ryder1965]. Finally, life course theories posit slower but still important changes over time as people advance through important transitions in their lives [@elder2016; @bardi2011].

Because all these theoretical perspectives have some empirical support, adjudicating when and how they apply is important for further understanding how early experiences and contemporary environments shape personal culture over the life course. That being said, scholars have struggled to reach a consensus on the relative importance of change during adulthood for explaining cultural differences. Over the span of a few years, most apparent changes in people’s responses to personal culture items appear to be transitory, with little evidence that adults make persistent changes of personal culture [@kiley2020; @vaisey2021]. When looked at over longer time horizons, however, there is at least some evidence that adults make persistent cultural changes, especially as they move through different social roles across the life course [@lersch2023].

These different conclusions are partly due to the fact that researchers decided to adjudicate the different perspectives by taking a “tournament of models” approach [@lersch2023: p. 228]. In this approach, researchers have formalized models of responses over time as demonstrating either “change” or “no change” and compared which model fits observed data better, functionally asking whether we ever observe within-person change in personal culture.

We argue, however, that asking only *whether* people change obscures what are actually highly compatible results. Whether change “really happened” in a given data set depends not only on sample populations, statistical methods and power, and item measurement, but crucially also on definitions of what counts as “change,” thus running the risk of getting caught up in unresolvable debates about related decisions. Moreover, it narrows the focus on particular inquiries such as, “do we see change in this measure?” Such inquiries are ill-suited to address what’s at the core of the theoretical debate. There is enough evidence to assume that neither socialization nor the environment alone can explain cultural differences. What’s left to answer is under what conditions, across which groups, and in what domains do we observe differences in the relative importance of intrapersonal change and interpersonal differences. This focus is necessarily comparative and requires an approach that quantifies forms of cultural change rather than calling victory for the one or the other perspective.

In this paper, we try to overcome previous debates and move away from the “tournament of models” approach toward a quantification of the processes that lead to differences in personal culture. Drawing on the seven panel data sets analyzed in these debates, we use established methods to quantify the amount of variance in responses that is due to between-people differences in baselines and the amount of variance in responses that is due to linear within-person change over time. This measure gives us a sense of the importance of intrapersonal change over the course of the panel relative to the differences that exist between people when they start, a quantity with more diverse applications than a simple judgment about whether change happened. We also explore features of questions and populations that explain differences in within-person change over time.

The pattern of results help clarify and, hopefully, transcend previous debates. We observe similar patterns across the data sets that have produced conflicting interpretations. In all data sets we see similar rates of individual-level change over time across a broad range of measures of personal culture. But quantifying these processes also reveals that intrapersonal change plays a substantially smaller role in explaining the variance in peoples’ personal culture than between-person differences in their baselines. Whether this amount of intrapersonal change is “meaningful” ultimately remains a theoretical question reflected in prior expectations concerning the question, population, and time. Nevertheless, the results suggest that the intrapersonal change observed (in some cases over more than 10 years) does not seem to play a substantial role in explaining personal cultural differences between adults.

While adding to existing debates, the quantification we propose extends them in new ways that enables scholars to more clearly adjudicate theoretical mechanisms at play. We provide an example by exploring differences in the relative importance of intrapersonal change for people with and without college degrees, comparing both within and across questions in ways that the tournament of models cannot.

>>>>

---
title: "Quantifying Stability and Change in Personal Culture Using Panel Data"
author:
- 
output:
  pdf_document:
    keep_tex: true
bibliography: kileybib.bib
csl: asa.csl
abstract: "Recent work has produced conflicting interpretations on the question of whether adults undergo durable changes in their personal culture over time. This paper asserts that this divergence is less due to their findings than the approach taken -- a tournament of models focusing on the presence or absence of change. To move beyond this, we reanalyze the data used previously in this debate with an approach that quantifies the processes of cultural difference. In doing so, we compare how much cultural difference is explained by intrapersonal change and interpersonal differences. Our results harmonize recent findings by showing that, while most measures of personal culture show change over time, the variance explained by intrapersonal change is typically small compared to interpersonal differences at baseline. Quantifying these processes offers new ways to address theoretical questions of when and how their relative importance shifts. We examplify this with the case of college experience."
fontsize: 11pt
---

@@

A recurring focus in the sociology of culture is the question of whether people, as they move through their lives and transition between social roles, change their personal culture –- the declarative and non-declarative attitudes, beliefs, values, practices, and dispositions [@lizardo2017; @kiley2020; @lersch2023].

This seemingly simple question underlies a number of central debates. For example, pragmatist theories of action claim that changes in social environments cause people to adapt their views and make new meanings [@swidler2001; @gross2009], while Bourdieusian practice theories argue that the “past conditions of production” leave a mark on people's personal culture that lasts throughout their lives [@bourdieu1990]. Similarly, models of social influence assume that people adapt their culture in the face of new information [@vaisey2010a], while the emphasis on cohort effects in models of aggregate social and cultural change requires them to be open to change while young but become fairly resistant to it as they age [@ryder1965]. Finally, life course theories posit slower but still important changes over time as people advance through important transitions in their lives [@elder2016; @bardi2011].

All of the above theoretical perspectives have some empirical support. All the more it is important to adjudicate when and how they apply to advance an understanding the differences in people's personal culture that we observe. That being said, scholars have struggled to reach a consensus on the relative importance of change during adulthood for explaining cultural differences. Over the span of a few years, most apparent changes in people’s responses to personal culture items appear to be transitory, with little evidence of persistent change among adults [@kiley2020; @vaisey2021]. When looked at over longer time horizons, however, there is at least some evidence that adults make persistent cultural changes, especially as they move through distinct social roles across the life course [@lersch2023].

These seemingly inconsistent findings are partly due to the fact that researchers have taken a “tournament of models” approach [@lersch2023: p. 228] in adjudicating the different perspectives. Accordingly, researchers have formalized models of responses to cultural items over time as demonstrating either “change” or “no change” to then compare which model fits observed data better, thus functionally asking whether we ever observe within-person change in personal culture.

We argue, however, that asking only *whether* people change obscures what are actually highly compatible results. Whether change “really happened” in a given data set depends not only on sample populations, item measurement, statistical methods and power, but crucially also on definitions of what counts as “change,” thus running the risk of getting caught up in unresolvable debates about related decisions. [MAYBE BETTER TO SAY HERE: WHICH QUICKLY RELEGATES THE POSSIBILITY OF AN ANSWER BACK TO AN ULTIMATELY UNRESOLVABLE THEORETICAL DEBATE.] Moreover, it narrows the focus on particular questions such as, “Do we see change in this measure?” Such inquiries are ill-suited to address what’s at the core of the theoretical debate. There is enough evidence to assume that neither socialization at an early age nor the influence of the contemporary environment alone can explain cultural differences. The question that remains is under what conditions, across which groups, and in what domains do we observe differences in the relative importance of intrapersonal change and interpersonal differences. This focus is necessarily comparative and requires an approach that quantifies forms of cultural change rather than calling victory for the one or the other perspective.

In this paper, we try to move beyond previous debates by moving away from the “tournament of models” approach toward a quantification of the processes that lead to differences in personal culture. Drawing on the seven panel data sets analyzed in these debates, we use established methods to quantify the amount of variance in responses that is due to between-people differences in baselines versus the amount of variance in responses that is due to linear within-person change over time. This measure gives us a sense of the importance of intrapersonal change over the course of the panel relative to the differences that exist between people when they start. We then explore features of questions and populations that explain differences in within-person change over time.

We observe similar patterns across the data sets that have produced conflicting interpretations. In all data sets we see similar rates of individual-level change over time across a broad range of measures of personal culture. But quantifying these processes also reveals that intrapersonal change plays a substantially smaller role in explaining the variance in peoples’ personal culture than between-person differences in their baselines. Of course, whether this amount of intrapersonal change is “meaningful” ultimately remains a theoretical question reflected in prior expectations concerning the question, population, and time. Nevertheless, the results suggest that the intrapersonal change observed (in some cases over more than 10 years) does not seem to play a substantial role in explaining personal cultural differences between adults.

This pattern of results helps clarify and, hopefully, transcend previous debates. What is more, the measure we propose offers a quantity with more diverse applications than a simple judgment about whether change happened. To examplify this, we apply the measure to explore differences in the relative importance of intrapersonal change for people with and without college degrees, comparing both within and across questions in ways that the tournament of models cannot.

While adding to existing debates, we hope the quantification we propose extends them in new ways and enables scholars to more clearly adjudicate theoretical mechanisms at play in shaping people's personal culture.
